\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2017-01-11 mSystems/rev/mSys_MAIN_rev.tex
New draft approach to rank stability islands, which leads to REVISION draft G
Changes to be committed:
	modified:   ../documentation/MediaInspector_macOSX_mSystems.jnlp
	modified:   figs/Fig4.eps
	modified:   figs/Fig5.eps
	deleted:    figs/_ed_supfig_corrank_IBS_C.pdf
	deleted:    figs/_ed_supfig_corrank_IBS_P1.pdf
	new file:   figs/supfig_corrank_HLS_StoolA_after.eps
	new file:   figs/supfig_corrank_HLS_StoolA_before.eps
	deleted:    figs/supfig_corrank_IBS_C.eps
	deleted:    figs/supfig_corrank_IBS_P1.eps
	modified:   mSys_MAIN_rev.pdf
	modified:   mSys_MAIN_rev.tex
	modified:   mSys_MandM_

% release codenumber (update it here!)
\newcommand\releasecode{REVISION draft F}

@@

%% Discussion
\bibitem {kilpatrick} {\bf Kilpatrick a M, Ives a R.} 2003. Species interactions can explain Taylor’s power law for ecological time series. Nature {\bf 422}:65–68.
\bibitem {ballantyne} {\bf Ballantyne IV F, J. Kerkhoff A.} 2007. The observed range for temporal mean-variance scaling exponents can be explained by reproductive correlation. Oikos {\bf 116}:174–180.
\bibitem {joao} {\bf Stein RR, Bucci V, Toussaint NC, Buffie CG, Rätsch G, Pamer EG, Sander C, Xavier JB.}  2013. Ecological modeling from time-series inference: insight into dynamics and stability of intestinal microbiota. PLoS Comput Biol {\bf 9}:e1003388.
\bibitem {mehta} {\bf Fisher CK, Mehta P.} 2014. Identifying keystone species in the human gut microbiome from metagenomic timeseries using sparse linear regression. PLoS One {\bf 9}:e102451.
\bibitem {bucci}  {\bf Bucci V, Tzen B, Li N, Simmons M, Tanoue T, Bogart E, Deng L, Yeliseyev V, Delaney ML, Liu Q, Olle B, Stein RR, Honda K, Bry L, Gerber GK.} 2016. MDSINE: Microbial Dynamical Systems INference Engine for microbiome time-series analyses. Genome Biol {\bf 17}:121.
\bibitem {koenig} {\bf Koenig JE, Spor A, Scalfone N, Fricker AD, Stombaugh J, Knight R, Angenent LT, Ley RE.} 2011. Succession of microbial consortia in the developing infant gut microbiome. Proc Natl Acad Sci {\bf 108}:4578–4585.
\bibitem {FD} {\bf Weber, J.} 1956. Fluctuation Dissipation Theorem. Phys. Rev. {\bf 101}:1620-6
\bibitem {sislands1} {\bf Ng SC, Lam EFC, Lam TTY, Chan Y, Law W, Tse PCH, Kamm M a, Sung JJY, Chan FKL, Wu JCY.} 2013. Effect of probiotic bacteria on the intestinal microbiota in irritable bowel syndrome. J Gastroenterol Hepatol {\bf 28}:1624–31.
\bibitem {sislands2} {\bf Carroll IM, Ringel-Kulka T, Siddle JP, Ringel Y.} 2012. Alterations in composition and diversity of the intestinal microbiota in patients with diarrhea-predominant irritable bowel syndrome. Neurogastroenterol Motil {\bf 24}:521-30
\bibitem {sislands3} {\bf Xu D, Gao J, Gillilland M, Wu X, Song I, Kao JY, Owyang C.} 2014. Rifaximin alters intestinal bacteria and prevents stress-induced gut inflammation and visceral hyperalgesia in rats. Gastroenterology {\bf 146}:484–496.e4.
\bibitem {sislands4} {\bf Rooks MG, Veiga P, Wardwell-scott LH, Tickle T, Segata N, Michaud M, Gallini CA, Beal C, van Hylckama-Vlieg JET, Ballal SA, Morgan XC, Glickman JN, Gevers D, Huttenhower C, Garrett WS, Hylckama-vlieg JET Van, Ballal SA, Morgan XC, Glickman JN, Gevers D, Huttenhower C, Garrett WS.} 2014. Gut microbiome composition and function in experimental colitis during active disease and treatment-induced remission. ISME J {\bf 8}:1403–17.
\bibitem {tikhonov} {\bf  Tikhonov M.} 2016. Community-level cohesion without cooperation. Elife {\bf 5}.

>>>>

% release codenumber (update it here!)
\newcommand\releasecode{REVISION draft G}

@@

%% Discussion
\bibitem {kilpatrick} {\bf Kilpatrick a M, Ives a R.} 2003. Species interactions can explain Taylor’s power law for ecological time series. Nature {\bf 422}:65–68.
\bibitem {ballantyne} {\bf Ballantyne IV F, J. Kerkhoff A.} 2007. The observed range for temporal mean-variance scaling exponents can be explained by reproductive correlation. Oikos {\bf 116}:174–180.
\bibitem {joao} {\bf Stein RR, Bucci V, Toussaint NC, Buffie CG, Rätsch G, Pamer EG, Sander C, Xavier JB.}  2013. Ecological modeling from time-series inference: insight into dynamics and stability of intestinal microbiota. PLoS Comput Biol {\bf 9}:e1003388.
\bibitem {mehta} {\bf Fisher CK, Mehta P.} 2014. Identifying keystone species in the human gut microbiome from metagenomic timeseries using sparse linear regression. PLoS One {\bf 9}:e102451.
\bibitem {bucci}  {\bf Bucci V, Tzen B, Li N, Simmons M, Tanoue T, Bogart E, Deng L, Yeliseyev V, Delaney ML, Liu Q, Olle B, Stein RR, Honda K, Bry L, Gerber GK.} 2016. MDSINE: Microbial Dynamical Systems INference Engine for microbiome time-series analyses. Genome Biol {\bf 17}:121.
\bibitem {koenig} {\bf Koenig JE, Spor A, Scalfone N, Fricker AD, Stombaugh J, Knight R, Angenent LT, Ley RE.} 2011. Succession of microbial consortia in the developing infant gut microbiome. Proc Natl Acad Sci {\bf 108}:4578–4585.
\bibitem {FD} {\bf Weber, J.} 1956. Fluctuation Dissipation Theorem. Phys. Rev. {\bf 101}:1620-6
%\bibitem {sislands1} {\bf Ng SC, Lam EFC, Lam TTY, Chan Y, Law W, Tse PCH, Kamm M a, Sung JJY, Chan FKL, Wu JCY.} 2013. Effect of probiotic bacteria on the intestinal microbiota in irritable bowel syndrome. J Gastroenterol Hepatol {\bf 28}:1624–31.
%\bibitem {sislands2} {\bf Carroll IM, Ringel-Kulka T, Siddle JP, Ringel Y.} 2012. Alterations in composition and diversity of the intestinal microbiota in patients with diarrhea-predominant irritable bowel syndrome. Neurogastroenterol Motil {\bf 24}:521-30
%\bibitem {sislands3} {\bf Xu D, Gao J, Gillilland M, Wu X, Song I, Kao JY, Owyang C.} 2014. Rifaximin alters intestinal bacteria and prevents stress-induced gut inflammation and visceral hyperalgesia in rats. Gastroenterology {\bf 146}:484–496.e4.
%\bibitem {sislands4} {\bf Rooks MG, Veiga P, Wardwell-scott LH, Tickle T, Segata N, Michaud M, Gallini CA, Beal C, van Hylckama-Vlieg JET, Ballal SA, Morgan XC, Glickman JN, Gevers D, Huttenhower C, Garrett WS, Hylckama-vlieg JET Van, Ballal SA, Morgan XC, Glickman JN, Gevers D, Huttenhower C, Garrett WS.} 2014. Gut microbiome composition and function in experimental colitis during active disease and treatment-induced remission. ISME J {\bf 8}:1403–17.
\bibitem {tikhonov} {\bf  Tikhonov M.} 2016. Community-level cohesion without cooperation. Elife {\bf 5}.

### 2017-01-11 mSystems/rev/mSys_MandM_rev.tex
New draft approach to rank stability islands, which leads to REVISION draft G
Changes to be committed:
	modified:   ../documentation/MediaInspector_macOSX_mSystems.jnlp
	modified:   figs/Fig4.eps
	modified:   figs/Fig5.eps
	deleted:    figs/_ed_supfig_corrank_IBS_C.pdf
	deleted:    figs/_ed_supfig_corrank_IBS_P1.pdf
	new file:   figs/supfig_corrank_HLS_StoolA_after.eps
	new file:   figs/supfig_corrank_HLS_StoolA_before.eps
	deleted:    figs/supfig_corrank_IBS_C.eps
	deleted:    figs/supfig_corrank_IBS_P1.eps
	modified:   mSys_MAIN_rev.pdf
	modified:   mSys_MAIN_rev.tex
	modified:   mSys_MandM_

The Rank Stability Index (RSI) is shown as a percentage in a separate bar on the right of the rank matrix plot in Figures \ref{fig:corrank_IBS_A} and \ref{fig:corrank_IBS_P2}. The RSI is strictly $1$ for an element whose range never changes over time, and is strictly $0$ for an element whose rank oscillates between the extremes over time. So, the RSI is calculated, per element, as $1$ less the quotient of the number of true rank hops taken between the number of maximum possible rank hops, all powered to $p$:
\begin{linenomath}
$${\rm RSI} = \left(1-\frac{\text{true rank hops}}{\text{possible rank hops}}\right)^p = \left(1-\frac{D}{(N-1)(t-1)}\right)^p$$
\end{linenomath}
where $D$ is the total of rank hops taken by the studied element, $N$ is the number of elements that have been ranked, and $t$ is the number of time samples. The power index $p=4$ was arbitrarily chosen to increase the resolution in the stable region.

>>>>

The Rank Stability Index (RSI) is shown as a percentage in a separate bar on the right of the rank matrix plot in Figures \ref{fig:corrank_HLS_abroad} and \ref{fig:corrank_HLS_returned} and Supplementary Figures \ref{supfig:corrank_HLS_before} and \ref{supfig:corrank_HLS_after}. The RSI is strictly $1$ for an element whose range never changes over time, and is strictly $0$ for an element whose rank oscillates between the extremes over time. So, the RSI is calculated, per element, as $1$ less the quotient of the number of true rank hops taken between the number of maximum possible rank hops, all powered to $p$:
\begin{linenomath}
$${\rm RSI} = \left(1-\frac{\text{true rank hops}}{\text{possible rank hops}}\right)^p = \left(1-\frac{D}{(N-1)(t-1)}\right)^p$$
\end{linenomath}
where $D$ is the total of rank hops taken by the studied element, $N$ is the number of elements that have been ranked, and $t$ is the number of time samples. The power index $p=4$ was arbitrarily chosen to increase the resolution in the stable region.

### 2017-01-11 mSystems/rev/mSys_discussion_rev.tex
New draft approach to rank stability islands, which leads to REVISION draft G
Changes to be committed:
	modified:   ../documentation/MediaInspector_macOSX_mSystems.jnlp
	modified:   figs/Fig4.eps
	modified:   figs/Fig5.eps
	deleted:    figs/_ed_supfig_corrank_IBS_C.pdf
	deleted:    figs/_ed_supfig_corrank_IBS_P1.pdf
	new file:   figs/supfig_corrank_HLS_StoolA_after.eps
	new file:   figs/supfig_corrank_HLS_StoolA_before.eps
	deleted:    figs/supfig_corrank_IBS_C.eps
	deleted:    figs/supfig_corrank_IBS_P1.eps
	modified:   mSys_MAIN_rev.pdf
	modified:   mSys_MAIN_rev.tex
	modified:   mSys_MandM_

Specifically, the analysis of the rank stability of the samples of healthy and IBS--diagnosed patients studied in our lab\cite{IBS}, suggests that the presence of \emph{rank stability islands} among medium-ranked taxa is a interesting feature. The higher stability of these taxa goes against the global meaning of the scaling index. Interestingly, this stability disappears when we look at IBS patients. From the literature, it seems that some genera from the families Comamonadaceae, Neisseriaceae and Carnobacteriaceae have been reported to lower their abundance in IBS patients against healthy controls \cite{sislands1}. In our case, we see that these families that are present in the rank stability island of the healthy patient decrease their rank stability index or even disappear in the IBS patient. However, we also see contradictory results in other families as Lactobacillaceae or Fusobacteriaceae, which seems to increase their abundance in IBS patients \cite{sislands2}, while we observe an increase of the former and a decrease of the latter. The presence of members of the Lactobacillae family have been reported to have positive effects against gut inflammation and visceral hyperalgesia \cite{sislands3}, usual symptoms of gastrointestinal disorders. It could be happening that a disorder in the stability of this particular group may help to arise the onset of symptoms associated to gastrointestinal disorders. The Aerococacceae family is also enriched in rats with IBS symptoms who have been treated with immunomodulators \cite{sislands4}. Inside this island of stability, we also have families as Fusobacteriaceae o Hallomonadaceae which include pathogenic genera in them, but that are not present in the IBS patient. It could be brought into question the rule of these taxa as key players in the phase transition of the microbiota, or whether they are more susceptible to perturbations than the most abundant. The types of interactions that could sustain this particular behavior are not clear, as these non-abundant taxa are not usually included in dynamic studies in order to obtain a community matrix. Further experiments and data analysis are needed to clarify whether this was a unique event, or whether it is a widespread feature of stable microbiotas.

>>>>

Specifically, the analysis of the rank stability of different periods of time belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife}, suggests that the presence of \emph{rank stability islands} among medium-ranked taxa is a interesting feature. Interestingly, this stability was compromised when the period was not an ordinary one, suggesting that those taxa was even sensitive to changes in the lifestyle. It could be brought into question the rule of these taxa as key players in the phase transition of the microbiota, or whether they are more susceptible to perturbations than the most abundant. The types of interactions that could sustain this particular behavior are not clear, as these non-abundant taxa are not usually included in dynamic studies in order to obtain a community matrix. Further experiments and data analysis are needed to clarify whether this was a unique event, or whether it is a widespread feature of stable microbiotas.

### 2017-01-11 mSystems/rev/mSys_results_rev.tex
New draft approach to rank stability islands, which leads to REVISION draft G
Changes to be committed:
	modified:   ../documentation/MediaInspector_macOSX_mSystems.jnlp
	modified:   figs/Fig4.eps
	modified:   figs/Fig5.eps
	deleted:    figs/_ed_supfig_corrank_IBS_C.pdf
	deleted:    figs/_ed_supfig_corrank_IBS_P1.pdf
	new file:   figs/supfig_corrank_HLS_StoolA_after.eps
	new file:   figs/supfig_corrank_HLS_StoolA_before.eps
	deleted:    figs/supfig_corrank_IBS_C.eps
	deleted:    figs/supfig_corrank_IBS_P1.eps
	modified:   mSys_MAIN_rev.pdf
	modified:   mSys_MAIN_rev.tex
	modified:   mSys_MandM_

The rank dynamics and stability plots in Figure \ref{fig:corrank_IBS_A} and \ref{fig:corrank_IBS_P2} show the variations in rank over time for the most dominant taxa and their calculated Rank Stability Index (RSI, as discussed in Material and Methods) for the taxa of a healthy subject (individual \emph{A}, Figure \ref{fig:corrank_IBS_A}) and for a subject diagnosed with IBS (patient \emph{P2}, Figure \ref{fig:corrank_IBS_P2}) of the IBS study\cite{IBS}. The taxa were listed ordered by their accumulated frequency over the time series, with the y-axis being the overall dominance axis for each sample set. Generally speaking, we observed that the most dominant taxa had the highest rank stability.

\begin{figure}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	%\includegraphics[width=0.99\textwidth]{results/corrank/IBS_h_A_amplicons_family_Rank.pdf}
	\includegraphics[width=0.99\textwidth]{figs/Fig4.eps}
	\caption{Rank variation over time for the most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods) for samples from a healthy subject (individual \emph{A}) studied in our lab\cite{IBS}.}
	\label{fig:corrank_IBS_A}
\end{figure}

\begin{figure}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	%\hspace*{-4.5mm}\includegraphics[width=0.99\textwidth]{results/corrank/IBS_P2_Metatranscriptores_Rank.pdf}
	\includegraphics[width=0.99\textwidth]{figs/Fig5.eps}
	\caption{Rank variation over time for the most dominant elements (taxa) and their calculated Rank Stability Index for samples from a subject diagnosed with irritable bowel syndrome (patient \emph{P2}) studied in our lab\cite{IBS}.}
	\label{fig:corrank_IBS_P2}
\end{figure}

Nevertheless, in the particular case of the healthy individual in Figure \ref{fig:corrank_IBS_A}, \emph{Burkholderiales} and \emph{Betaproteobacteria} (the 18th and 25th taxa in the dominance axis) comparatively showed very low rank stability regarding similar dominant taxa while, on the other hand, \emph{Comamonadaceae}, \emph{Lactobacillaceae}, \emph{Fusobacteriaceae}, \emph{Aerococcaceae} and \emph{Carnobacteriaceae} showed higher stability than other more dominant taxa, forming a kind of \emph{rank stability island} for medium--ranked taxa around position 40 in the dominance axis, and thus colored brigher, since they show a moderately stable (RSI over 75\%).

In the IBS--diagnosed patient in Figure \ref{fig:corrank_IBS_P2}, beyond the differences in dominance for the particular taxa, we still observed that the most dominant were the most rank stable. However, as opposed to the healthy individual results, far from presenting a \emph{rank stability island}, the medium-ranked taxa were very rank unstable, mostly due to transient (often one or two consecutive samples) yet dramatic drops in their relative abundance, which usually occurred more than twice during their time series. For example, this was the case of \emph{Sphingobacteriales} with two non-consecutive samples dropping in the ranking to 111th. In other cases, the high rank instability came from a rank fluctuation throughout the time series, such as for \emph{Streptococcaceae} and \emph{Burkholderiales}, which ranked 26th and 29th, respectively, in the overall dominance axis but had a very low RSI.

We also found a \emph{rank stability island} for medium-ranked taxa in the other healthy subjects (\emph{B} and \emph{C}) in the IBS study\cite{IBS} together with its total absence for the other IBS--diagnosed patient (patient \emph{P1}), which also showed very high rank instability in its medium-ranked taxa. See Supplementary Figure \ref{supfig:corrank_IBS_C} and Supplementary Figure \ref{supfig:corrank_IBS_P1} for further examples of rank plots for a healthy and mild IBS--diagnosed patients, respectively.

\begin{supfig}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	\includegraphics[width=0.99\textwidth]{figs/supfig_corrank_IBS_C.eps}
	\caption{Rank variation over time for the most dominant elements (taxa) and their calculated RSI for samples from another healthy subject studied in our lab\cite{IBS}.}
	\label{supfig:corrank_IBS_C}
\end{supfig}

\begin{supfig}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	%\hspace*{-4.5mm}
	\includegraphics[width=0.99\textwidth]{figs/supfig_corrank_IBS_P1.eps}
	\caption{Rank variation over time for the most dominant elements (taxa) and their calculated RSI for samples from a subject diagnosed with mild IBS and studied in our lab\cite{IBS}.}
	\label{supfig:corrank_IBS_P1}
\end{supfig}

>>>>

The rank dynamics and stability plots in Figure \ref{fig:corrank_HLS_abroad} and \ref{fig:corrank_HLS_returned} show the variations in rank over time for the most dominant taxa and their calculated Rank Stability Index (RSI, as discussed in Material and Methods) for the gut microbiome taxa of a healthy subject, the individual \emph{A} in the host lifestyle study \cite{hostlife}. The Figure \ref{fig:corrank_HLS_abroad} covers the period when the individual is travelling abroad and the Figure \ref{fig:corrank_HLS_returned} covers the subsequent period. The taxa were listed ordered by their accumulated frequency over the time series, with the y-axis being the overall dominance axis for each sample set. Generally speaking, we observed that the most dominant taxa had the highest rank stability.

\begin{figure}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	\includegraphics[width=1.0\textwidth]{figs/Fig4.eps}
	\caption{Rank variation over time for the 50 most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods) for a special period (days 72 to 122, travelling abroad) belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife}.}
	\label{fig:corrank_HLS_abroad}
\end{figure}

\begin{figure}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures \hspace*{-4.5mm}
	\includegraphics[width=1.0\textwidth]{figs/Fig5.eps}
	\caption{Rank variation over time for the 50 most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods) for an ordinary period (days 123 to 256, after the trip) belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife}.}
	\label{fig:corrank_HLS_returned}
\end{figure}

For the trip abroad in Figure \ref{fig:corrank_HLS_abroad}, beyond the differences in dominance for the particular taxa, we still observed that the most dominant were the most rank stable. Moreover, the medium-ranked taxa were quite rank unstable, mostly due to transient (often one or two consecutive samples) yet dramatic drops in their relative abundance, which usually occurred more than twice during their time series.

Nevertheless, in the particular case of the next period, the one subsequent to the trip, in Figure \ref{fig:corrank_HLS_returned}, some taxa showed higher stability than other more dominant taxa, forming a kind of \emph{rank stability islands} for medium--ranked taxa, since they show a moderately stable index (RSI roughly over 70\%). In particular, this is the case for the genera \emph{Actinomyces}, \emph{Leuconostoc}, \emph{Lachnobacterium}, \emph{Eggerthella}, \emph{Clostridium} and	\emph{Collinsella}. For those genera, both the overall rank and the RSI were clearly lower during the trip (RSI under 70\%). \emph{Actinomyces} and \emph{Lachnobacterium} are even not shown in Figure \ref{fig:corrank_HLS_abroad} because they sank to positions 56 and 77, respectively. On the other hand, \emph{Leuconostoc} was the less sensitive to the change of lifestyle.

We also found those \emph{rank stability islands} for medium-ranked taxa in the other periods belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife} (see Supplementary Figure \ref{supfig:corrank_HLS_before} and Supplementary Figure \ref{supfig:corrank_HLS_after} for the corresponding rank plots).

\begin{supfig}
	\centering
	\includegraphics[width=1.0\textwidth]{figs/supfig_corrank_HLS_StoolA_before.eps}
	\caption{Rank variation over time for the 50 most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods) for an ordinary period (days 0 to 70, before the trip) belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife}.}
	\label{supfig:corrank_HLS_before}
\end{supfig}

\begin{supfig}
	\centering
	\includegraphics[width=1.0\textwidth]{figs/supfig_corrank_HLS_StoolA_after.eps}
	\caption{Rank variation over time for the 50 most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods) for an ordinary period (days 257 to 364, further after the trip) belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife}.}
	\label{supfig:corrank_HLS_after}
\end{supfig}
